Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-05 paper/paper.tex
software

\author[0000-0003-0872-7098]{Adrian M. Price-Whelan}
\affil{Department of Astrophysical Sciences, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.2$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This data-driven model contains no knowledge of stellar astrophysics or the
Milky Way, but rather derives its precision from its generative model of noisy
parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The \xd-optimized prior is used to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a posterior parallax and distance sampling for 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

\keywords{
  catalogs
  ---
  Hertzsprung--Russell and C--M diagrams
  ---
  methods: statistical
  ---
  parallaxes
  ---
  stars: distances
}

@@

The \gaia\ Mission (\citealt{gaia16}) will soon deliver more than a billion
stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be both precise and purely astrometric.
\gaia\ will use astrometric parallaxes to determine the distances of the more
precise stars, and then calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
parallax estimates to less precise stars.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

However, the use of physical models for distance estimation is not necessary.
It is possible to build a stellar photometric model from the data themselves,
because there are stars at all luminosities with useful parallax information.
One challenge is that different stars are observed at different levels of parallax precision,
so it requires relatively sophisticated technology to build this data-driven model
using all of the data available, fairly and responsibly.

@@

We then use the novel \xd\ output---the deconvolved \cmd---as
a prior for use in inference of individual-star parallaxes.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is technically using the data twice and is therefore only approximately valid.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied.\footnote{For a primer on ``empirical Bayes'', see \url{https://en.wikipedia.org/wiki/Empirical_Bayes_method}.}

A probabilistically valid approach that wouldn't re-use the data is to perform
a full hierarchical Bayesian inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to a related
project \citep{leistedtHogg2017}, in which we use a slightly less appropriate
model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

This project is fundamentally a demonstration of concept:
We are only using a subset of the ``small'' (relative to the expected final \gaia\ data set)
\tgas\ Catalog.
We are performing only an approximation to full Bayesian hierarchical inference.
We have to use photometry that is ground-based, rather than the full-precision,
space-based photometry that the \gaia\ Mission will deliver.
However, as we show below, we get very good results in terms of parallax precision.
It is promising that we will be able to obtain even better results
using later \gaia\ data releases, and eventually infer distances for $>10^9$
\gaia\ stars using photometric parallax methodologies but without any
commitment to---or even use of---physical models of stars or the Milky Way.

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{toy.png}
\caption{ {\bf Why does this work}: An example of generating posterior beliefs of true values from noisy data by using the noise-deconvolved distribution of data as a prior. The upper-left panel visualizes the simplified toy distribution. The red lines are the $1\sigma$ contours of the distribution, and the red points are 1024 samples from the distribution. These same $1\sigma$ contours are drawn in the bottom left panel. In the upper-right and lower-left panel, the grey points are these same 1024 samples of the distribution but with some measurement uncertainty added to them. The error bars show the $1\sigma$ measurement uncertainties. The black points are 5 randomly chosen data to highlight the method. In the lower-left panel, the blue lines represent the noise deconvolved, best estimate for the underlying distribution. The blue points represent the posterior belief of the true values of the noisy black points using the noise deconcolved distribution as the prior. The red points represent the actual true values. The lower-right panel shows the posterior belief of the true value for all the points. They are colored by the measurement uncertainty, showing that data with larger measured variance are more influenced by the prior and lie towards the center of the noise deconvolved distribution, or prior. In contrast, more certain measurements tend to lie closer to their measured value. The blue points are better than the black for doing stuff.}
\label{fig:toy}
\end{figure}

In this paper we use the noise-deconvolved \cmd\ as a prior to infer more precise parallax probability distribution functions (pdfs) from \gaia\ measurements.
Understanding the validity of this method can be challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we demonstrate the generaly methodology using a
simpler toy model with simulated data.
Instead of inferring parallax pdfs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer posterior pdfs of the value $Y$.
We perform this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the y-direction, with a running mean $\mu = m\,x + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness or intrinsic scatter of the \cmd\ due to age, mass, and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in \figurename~\ref{fig:toy}, upper-left panel.
Similarly, noisy samples from this distribution are shown in the upper-right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample from the distribution, from $y_n$, a sample from the noise distribution.

Setting up the problem in more detail, we have the true samples from the underlying distribution and the noisy samples
\begin{eqnarray}
y_{\true, n} \, &=& \, m_{\true}\,x_n + b_{\true} + \Delta_n \\
y_n \, &=& \, y_{\true,n} + \epsilon_n \quad,
\label{eq:ytrue}
\end{eqnarray}
where $\Delta_n$ is drawn from the Gaussian function $\mathcal{N}(\Delta_n \given 0, t^2)$---which accounts for the intrinsic width---and $\epsilon_n$ is drawn from the Gaussian function $\mathcal{N}(\epsilon_n \given 0, \sigma_n^2)$---the noise distribution---so $\sigma_n$ is the measurement uncertainty.
These represent true and noisy draws from a distribution that is a straight line with some thickness $t$, and are visualized in the upper panels of \figurename~\ref{fig:toy}.
The upper-left shows 1024 realizations of the true values, and the upper-right, as well as lower-left, show 1024 realizations of the noisy values $y_n$ and their associated measurement uncertainties $\sigma_n$ as the grey points; we highlight 5 random points in black.

To infer a posterior pdf of the true $y$ value for each measured $y$ value we need a prior. We use an estimate of the underlying true distribution that the noisy data $y_n$ are drawn from, which we build from the noisy $y$ data themselves.
We learn the noise-deconvolved function using marginalized, maximum-likelihood estimation (MMLE), which marginalizes over the true data values and maximizes the likelihood of the data. Here our likelihood for each individual datum is
\begin{eqnarray}
p(y_n \given m, b, t^2, \sigma_n^2) \ &=& \ \int \mathcal{N}(y_n \given y_{\true,n}, \sigma_n^2) \, \mathcal{N}(y_{\true,n} \given m\,x_n + b, t^2 ) \, \mathrm{d}y_{\true, n} \\
&=&  \ \mathcal{N}(y_n \given m\,x_n + b, t^2 + \sigma_n^2) \quad,
\label{eq:toyLike}
\end{eqnarray}
where we have marginalized over the true sample $y_{\true, n}$. The resulting likelihood is a Gaussian because the noise model and our assumptions about the intrinsic width are both Gaussian. The likelihood of the full data set $\{y_n\}$ is then the product of the individual likelihoods
\begin{equation}
p(\{y_n\} \given m, b, t^2, \sigma_n^2) = \prod_n\, p(y_n \given m, b, t^2, \sigma_n^2) \quad .
\label{eq:toyLikeFull}
\end{equation}
Maximizing this joint likelihood of all the data is the same as minimizing
\begin{equation}
\chi^2 = \sum_n \frac{[y_n - [m\,x_n + b]]^2}{[\sigma_n^2 + t_n^2]} + \sum_n\,\ln(2\pi\,[\sigma_n^2 + t^2])
\quad .
\label{eq:chisq}
\end{equation}

The MMLE gives us the best fit parameters $\hat{m}$, $\hat{b}$, and $\hat{t}$. In \figurename~\ref{fig:toy}, lower-left panel, the true functions $y_{\true} = m_{\true}\,x + b_{\true} \pm t_{\true}$ are shown as the red lines, and the best fit MMLE functions $\hat{y}_{\true} = \hat{m}\, x + \hat{b} \pm \hat{t}$ are shown as the blue lines. Now that we have an estimate of the true underlying distribution $\hat{y}_{\true}$, we use this as a prior to infer a posterior belief of the true values for each datum $p(y_{\true,n} \given y_n, \tilde{\sigma}_n^2)$.

Here is a good point to step back and acknowledge that our method is a
bit strange. We are using the data to determine the
noise-deconvolved distribution (an estimate of the true
distribution), and then using that noise-deconvolved distribution to
make inferences about the data. We are technically using the data
twice! To be fully, technically correct, we should \emph{either} learn the true
distribution while leaving one datum out, the one datum we want to
infer the true y value for, or else perform a full hierarchical Bayesian inference,
in which we treat the \cmd\ parameters probabilistically along with everything else.
With large data sets, the MMLE or Empirical Bayes approximation is safe,
in some sense because no individual data point makes a large difference to
the inference.

Returning to the inference:
We use Bayes' theorem to turn the likelihood of the noisy data into a posterior belief over the true $y$ value:
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
\mathrm{where}\;p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
\mathrm{and}\;p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}
In \figurename~\ref{fig:toy} lower-left panel, the blue points represent the posterior beliefs of the true values $y_{\true,n}$ for the 5 highlighted black points from the upper-right panel. The red points are the associated true $y$ values.
In the lower-right panel of \figurename~\ref{fig:toy}, the blue points show the expectation value and $1\sigma$ uncertainty of the posterior belief for all 1024 $y_n$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance, $t^2$, and the measurement variance, $\sigma_n^2$.
The points are colored by their measurement variance $\sigma_n^2$, which shows that noisier measurements are more influenced by the prior and get pulled closer to the center of the MMLE distribution, our prior. In contrast, more certain measurements remain closer to their measured values, $y_n$. We now have better estimates of each true y value by inferring each true y value from the measured y value and using an estimate of the true distribution the measured y value was drawn from as the prior.

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical Bayes, or maximum-marginalized likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \panstarrs-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

\subsection{Setting up the Inference}

@@

Equation~\ref{eq:bayes} does not fully specify this procedure.
Because our prior comes from the \cmd, our inference will include not only the parallax from \gaia, but also the photometry of the stars. Specifically, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$.

>>>>

\author[0000-0003-0872-7098]{Adrian Price-Whelan}
\affil{Princeton University Observatory, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

\begin{abstract}\noindent % trust me
Conversion of a noisy parallax measurement into a posterior
belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for a large fraction of the
\gaia\ \tgas\ Catalog from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information for $1.2$ million \tgas\ stars
from \tmass\ to build a noise-deconvolved empirical prior distribution for stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a
transformed space where the measurement uncertainty is close to Gaussian.
The \xd-optimized prior is used to perform parallax inferences for every star, yielding a precise photometric parallax estimate and uncertainty (and full posterior) for every star.
Our posterior photometric parallax estimates are more precise than the
\tgas\ catalog entries by a median factor of 1.2, and 14 percent
are more precise by a factor $>2$. They are also more precise than
previous examples of Bayesian distance estimates using a spatial prior.
We independently validate our parallax inferences distances by looking
at members of Milky Way star cluster M67, which is not visible as a
cluster in the \tgas\ parallax estimates, but appears clearly as a
cluster in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for 1.4 million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

\keywords{
  catalogs
  ---
  Hertzsprung--Russell and C--M diagrams
  ---
  methods: statistical
  ---
  parallaxes
  ---
  stars: distances
  ---
  stars: statistics
}

@@

The \gaia\ Mission (\citealt{gaia16}) will deliver more than a billion stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be both precise and purely astrometric.
\gaia\ will use astrometric parallax to determine the distances of the more precise
stars, and then calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
parallax estimates to less precise stars.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

However, the use of physical models for distance estimation is not necessary.
It is possible to build a stellar photometric model from the data themselves,
because there are stars at all luminosities with useful parallax information,.
One challenge is that different stars are observed at different levels of parallax precision,
so it requires relatively sophisticated technology to build this data-driven model
using all the data available, fairly and responsibly.
Luckily, we have this technology!

@@

We then use the novel \xd\ output---the deconvolved \cmd---as
a prior for use in inference of individual-star parallaxes.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is technically using the data twice and is therefore only approximately valid.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied\footnote{For a start, try \url{https://en.wikipedia.org/wiki/Empirical_Bayes_method}.}.

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to a related
project \citep{leistedtHogg2017}, in which we use a slightly less appropriate
model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

This project is fundamentally a demonstration of concept:
We are only using a subset of the ``small'' (relative to the expected final \gaia\ data set)
\tgas\ Catalog.
We are performing only an approximation to full Bayesian hierarchical inference.
We have to use photometry that is ground-based, rather than the full-precision,
space-based photometry that the \gaia\ Mission will deliver.
However, as we show below, we get very good results in terms of parallax precision.
It is promising that we will be able to obtain even better results
in later \gaia\ data releases, and eventually put distances on $>10^9$
\gaia\ stars using photometric parallax methodologies but without any
commitment to---or even use of---physical models of stars or the Milky Way.

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{toy.png}
\caption{ {\bf Why does this work}: An example of generating posterior beliefs of true values from noisy data by using the noise-deconvolved distribution of data as a prior. The upper left panel visualizes the simplified toy distribution. The red lines are the $1\sigma$ contours of the distribution, and the red points are 1024 samples from the distribution. These same $1\sigma$ contours are drawn in the bottom left panel. In the upper right and lower left panel, the grey points are these same 1024 samples of the distribution but with some measurement uncertainty added to them. The error bars show the $1\sigma$ measurement uncertainties. The black points are 5 randomly chosen data to highlight the method. In the lower left panel, the blue lines represent the noise deconvolved, best estimate for the underlying distribution. The blue points represent the posterior belief of the true values of the noisy black points using the noise deconcolved distribution as the prior. The red points represent the actual true values. The lower right panel shows the posterior belief of the true value for all the points. They are colored by the measurement uncertainty, showing that data with larger measured variance are more influenced by the prior and lie towards the center of the noise deconvolved distribution, or prior. In contrast, more certain measurements tend to lie closer to their measured value. The blue points are better than the black for doing stuff.}
\label{fig:toy}
\end{figure}

In this paper we use the noise deconvolved observed \cmd\ as a prior to infer more precise parallax pdfs from \gaia\ measurements.
Understanding this method is challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we will demonstrate this method with a simpler toy model.
Instead of inferring more precise parallax pdfs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer more precise pdfs of the value $Y$.
We will do this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the y-direction, with a running mean $\mu = mx + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness of the \cmd\ due to age, mass and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in Figure \ref{fig:toy}, upper left panel.
Similarly, noisy samples from this distribution are shown in the upper right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample of the distribution, from $y_n$, a noisy sample of the distribution.

Setting up the problem in more detail, we have the true samples from the underlying distribution and the noisy samples
\begin{eqnarray}
y_{\true, n} \, &=& \, m_{\true}\,x_n + b_{\true} + \Delta_n \\
y_n \, &=& \, y_{\true,n} + \epsilon_n \quad,
\label{eq:ytrue}
\end{eqnarray}
where $\Delta_n$ is drawn from the Gaussian function $\mathcal{N}(\Delta_n \given 0, t^2)$, and $\epsilon_n$ is drawn from the Gaussian function $\mathcal{N}(\epsilon_n \given 0, \sigma_n^2)$, so $\sigma_n$ is the measurement uncertainty.
These represent true and noisy draws from a distribution that is a straight line with some thickness $t$, and are visualized in the upper panels of Figure \ref{fig:toy}
The upper left shows 1024 realizations of the true values, and the upper right, as well as lower left, show 1024 realizations of the noisy values $y_n$ and their associated measurement uncertainties $\sigma_n$ as the grey points. Here we highlight 5 random points in black to highlight the method, but the same method is being applied to all the data.

To infer a pdf of the true y value for each measured y value we need a prior. For our prior we will use an estimate of the underlying true distribution that the noisy data $y_n$ are drawn from, which we build from the noisy y data themselves.
We learn the noise-deconvolved function using marginalized maximum likelihood estimation (MML), which marginalizes over the true data values (we never need to know them) and maximizes the likelihood of the data. Here our likelihood for each individual datum is
\begin{eqnarray}
p(y_n \given m, b, t^2, \sigma_n^2) \ &=& \ \int \mathcal{N}(y_n \given y_{\true,n}, \sigma_n^2) \, \mathcal{N}(y_{\true,n} \given m\,x_n + b, t^2 ) \, \mathrm{d}y_{\true, n} \\
&=&  \ \mathcal{N}(y_n \given m\,x_n + b, t^2 + \sigma_n^2) \quad,
\label{eq:toyLike}
\end{eqnarray}
where we have marginalized over the true sample $y_{\true, n}$. The resulting likelihood is a Gaussian because the noise model is a Gaussian. The likelihood of the full data set is then the product of the individual likelihoods
\begin{equation}
\prod_n\, p(y_n \given m, b, t^2, \sigma_n^2) \quad .
\label{eq:toyLikeFull}
\end{equation}
Maximizing this joint likelihood of all the data is the same as minimizing
\begin{equation}
\chi^2 = \sum_n \frac{[y_n - [m\,x_n + b]]^2}{[\sigma_n^2 + t_n^2]} + \sum_n\,\ln(2\pi\,[\sigma_n^2 + t^2])
\quad .
\label{eq:chisq}
\end{equation}

This MML gives us the best fit parameters $\hat{m}$, $\hat{b}$, and $\hat{t}$. In Figure \ref{fig:toy}, lower left panel, the true functions $y_{\true} = m_{\true}\,x + b_{\true} \pm t_{\true}$ are shown as the red lines, and the best fit MML functions $\hat{y}_{\true} = \hat{m}\, x + \hat{b} \pm \hat{t}$ are shown as the blue lines. Now that we have an estimate of the true underlying distribution $\hat{y}_{\true}$, we use this as a prior to infer a posterior belief of the true values for each datum $p(y_{\true,n} \given y_n, \tilde{\sigma}_n^2)$.

Here is a good point to step back and acknowledge that our method is a
bit strange. We are using the set of data to determine the
noise-deconvolved distribution, which is an estimate of the true
distribution, and then using that noise-deconvolved distribution to
make inferences on the data. We are technically using the data
twice! To be fully, technically correct, we should \emph{either} learn the true
distribution while leaving one datum out, the one datum we want to
infer the true y value for, or else go to full hierarchical Bayesian inference,
in which we treat the \cmd\ parameters probabilistically along with everything else.
With large data sets, the MML or Empirical Bayes approximation is safe,
in some sense because no individual data point makes a large difference to
the inference.

Returning the inference:
We use Bayes theorem to turn the likelihood of the noisy data into a posterior believe over the true y value.
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
\mathrm{where}\;p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
\mathrm{and}\;p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}
Since everything is Gaussian, this can be done analytically by multiplying Gaussians together. In Figure \ref{fig:toy} lower left panel, the blue points represent the posterior belief of the true value $p(y_{\true,n} \given y_n, \tilde{\sigma}_n)$ for the highlighted black points. The red points are the associated true y values.
In Figure \ref{fig:toy}, lower right panel, the blue points show the expectation value and $1\sigma$ uncertainty of the posterior belief for all 1024 $y_n$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance, $t^2$, and the measurement variance, $\sigma_n^2$.
The points are colored by their measurement variance $\sigma_n^2$, which shows that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior. In contrast, more certain measurements remain closer to their measured values, $y_n$. We now have better estimates of each true y value by inferring each true y value from the measured y value and using an estimate of the true distribution the measured y value was drawn from as the prior.

@@

To infer the true parallax to a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical Bayes, or maximum-marginalized likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar, for example, a white dwarf, inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \panstarrs-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

\subsection{Setting up the Inference}
 \emph{Think about presentation, prioritizing/highlighting prior but logically building it from an full inference perspective and needing a prior, then dive into prior stuff. It's confusing right now jumping between inference and prior building !!}

@@

Equation~\ref{eq:bayes} is too simplistic to describe this procedure.
Because our prior is the \cmd, our inference will include not only the parallax from \gaia, but also the photometry of the stars. Specifically, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$.

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\mathbf{y}$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{0.2J^c_n} = 10^{0.2\,M^c_{J,n} + 2}
\label{eq:transform}
\end{equation}
The transformation of the dust corrected absolute magnitude is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain. Gaussian uncertainties are required to generate our prior using \xd, but more about this and our dust corrections below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\mathbf{Y}_n = [10^{0.2\,M^{true}_{J,n} + 2}, (J-K)^{true}_n] \\
\mathrm{data \, vector} \quad &\mathbf{y}_n = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

We assume that the parallax likelihood is Gaussian, as well as the $(J-K)^c_n$ color likelihood. So our full likelihood of the $n$th star is
\begin{equation}
p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) = \mathcal{N}(\mathbf{y}_n \given \mathbf{Y}_n, V_n)
\end{equation}
with \\
\[
V_n = \begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & (\sigma_{\varpi}\,10^{0.2\,J_n^C})^2
\end{bmatrix}
\]
where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$ band uncertainty, and therefore the dominant uncertainty in the transformed absolute magnitude. So we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty.

We can now write the multidimensional analogous to Equation \ref{eq:bayes},
\begin{equation}
p(\mathbf{Y}_n \given \mathbf{y}_n, I_n) = p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) \, p(\mathbf{Y}_n) \label{eq:posterior}
\end{equation}
where $p(\mathbf{y}_n \given \mathbf{Y}_n, I_n)$ is the likelihood of the data vector, and $p(\mathbf{Y}_n)$ is our prior.
Before we describe how this can be turned into a posterior belief on the true parallax, we turn our attention to the prior, which will be generated from the data themselves. We will use XD, which will take advantage of the Gaussian likelihood.
